---
title: "Assignment 5: Data Visualization"
author: "Brian Mulu Mutua"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Initial setup
#Checking working directory 
getwd()

#Loading necessary libraries
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

#Reading processed data files
processedNTL.LTER.data <- read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"))
processedNiwot.Ridge.data <- read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"))

#2 Making sure R is reading dates as date format
#Checking date format of the loaded data for the North Temperate Lakes LTER Data
glimpse(processedNTL.LTER.data)
#Adjusting date format using lubridate
processedNTL.LTER.data$sampledate <- ymd(processedNTL.LTER.data$sampledate)
#Using class function to show that date format has been updated successfully
class(processedNTL.LTER.data$sampledate)


#Checking date format of the loaded data for the Niwot Ridge Long-Term Ecological Research (LTER) station
glimpse(processedNiwot.Ridge.data)
processedNiwot.Ridge.data$collectDate <- ymd(processedNiwot.Ridge.data$collectDate)
#Using class function to show that date format has been updated successfully
class(processedNiwot.Ridge.data$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Building personal theme
brian_theme_A05 <- theme_base()  + 
  theme(
    line =  element_line(colour = "indianred"),
    #rect =              element_rect(),
    #text =              element_text(),
    
    # Modified inheritance structure of text element
    plot.title = element_text(family = "Helvetica-Narrow",
                              face = "bold",
                              size = 20),
    axis.title.x = element_text(family="Helvetica-Narrow"),
    axis.title.y = element_text(family="Helvetica-Narrow"),
    axis.text = element_text(family="Helvetica-Narrow"),
    
    # Modified inheritance structure of line element
    axis.ticks = element_line(colour = "indianred",linewidth = 0.5),
    #panel.grid.major =  element_line(),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    plot.background = element_rect(colour = "bisque1"),
    #panel.background =  element_rect(),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    legend.position = "bottom",
    legend.background = element_rect(color = "bisque3")
    
    #complete = TRUE
    )

#Setting personal theme as default theme
theme_set(brian_theme_A05)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 Setting up plot
NTL.LTER.Question4Plot <- processedNTL.LTER.data %>% 
  ggplot(aes(x=tp_ug,y=po4,color=lakename)) + geom_point() + xlim(0,150) + ylim(0,50) + geom_smooth(method=lm) + brian_theme_A05
NTL.LTER.Question4Plot

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: