---
title: "Assignment 2: Coding Basics"
author: "Brian Mulu Mutua"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Using sequence function assigned the name, 'thirty_sequence' to generate the set of numbers between 1 and 30, increasing by threes

ten_number_sequence <- seq(1,30,3)

#2. Computing the mean and median of the sequence 

ten_number_sequence_mean <- mean(ten_number_sequence)
ten_number_sequence_median <- median(ten_number_sequence)

#3. Checking if mean of the sequence is greater than the median

ten_number_sequence_mean > ten_number_sequence_median

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

#Approach 1 
```{r basics 2}
#5&#6. Creating a  series of vectors of student names, test scores, and pass/fail status and labeling the vectors accordingly in the comments

student_names1 <- c("George Costanza", "Elaine Benes", "Jerry Seinfeld","Cosmo Kramer") #Vector of type character
student_test_scores <- c(45,70,30,60) #Vector of type integer
student_pass_status <- c(FALSE,TRUE,FALSE,TRUE) #Vector of type logical

#7. Combining each of the vectors in a data frame
df_student_names_and_test_scores <- data.frame(student_names1,student_test_scores,student_pass_status) #Creating data frame of student names and test scores

#8. Labeling the columns of the data frame with informative titles
colnames(df_student_names_and_test_scores) <- c("StudentName","StudentTestScore","HasStudentPassed")
print(df_student_names_and_test_scores)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: In a matrix, all columns must have the same mode and length. A data frame is more general than a matrix and consists of a list of vectors of equal length, which could have different modes. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#10. Creating a function with an if/else statement to check the vector of test scores and check if the given test score is a passing grade or not

has_student_passed_test <- function(test_scores){
  for(each_score in test_scores){
    pass_status <- c(student_pass_status,ifelse(each_score >= 50,TRUE,FALSE))
  }
  print(test_scores)
  print(pass_status)
}
has_student_passed_test(student_test_scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: If the student test score was greater or equal to 50, then the "TRUE" condition worked and if the value was less than 50, then the "FALSE" condition worked. 
